---
title: "Example Rmd"
author: "ipumsi"
date: 'July 27th 2022'
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Read in my_data (.csv)

Base R can handle .csv and .txt files "out of the box", but these can have limitations. **Note:** With Rmd, comments/text do not require the "#" comment symbol. Instead we specificy where the `code` will go using "code chunks", as below. 

```{r}

my_data <- read.csv("example_data.csv")

colnames(my_data)
table(my_data$SEX)
table(my_data$EDATTAIN)
table(my_data$EMPSTAT)
```


To read in different file types, we will need to download R packages. 

## Install Packages

R packages are add-ons that expand the functionality of base R. Its helpful  to think of R as a series of books. "Base R" is the starter set of "books" needed for R to function. install.packages() downloads new the "book" and adds it to your collection. This step needs to be done only once, or periodically to update. Note: quotation marks ("") ARE REQUIRED to specify a package name. openxlsx is a very useful package that allows read/write of xls/xlsx files. haven provides functions to read from various stats packages, including SPSS, STATA, and SAS. Notably, haven is able to read in value and value labels

```{r}
if(!require("openxlsx")) {
  install.packages("openxlsx")
}

if (!(require("haven"))) {
  install.packages("haven")
}
```

## Load Packages

In order to load the package, we use library(). This is like taking the book off your shelf and opening it to read/use. This step needs to be done each time you open R/Rstudio. so it's helpful to have a section at the top of the script that loads all packages needed. Note: quotation marks ("") are NOT required for the library function.

```{r}
library(haven)
library(openxlsx)
```

## Read in my_data (.xlsx)

Excel can be convenient, but like .csv files does not allow for labeled values, like many other stats programs.

```{r}
my_data <- read.xlsx("example_data.xlsx")

colnames(my_data)
table(my_data$SEX)
table(my_data$EDATTAIN)
table(my_data$EMPSTAT)

```
## Read in my_data (with Labels)
```{r}
my_data <- read_dta("example_data.dta")
```


*See Environment Panel:* haven labels are complex; factors are more simple.

```{r}
my_data$SEX <- as_factor(my_data$SEX)
my_data$EDATTAIN <- as_factor(my_data$EDATTAIN)
my_data$EMPSTAT <- as_factor(my_data$EMPSTAT)
```


#### one way tables for factors

```{r}
table(my_data$SEX)
table(my_data$EDATTAIN)
table(my_data$EMPSTAT)
```

#### or the range for numerics

```{r}
range(my_data$AGE)
```


## Crosstabs (two way tables)
```{r}
table(my_data$EMPSTAT, my_data$SEX)
```

#### Subset by a value - those too young to work

```{r}
sub_data <- my_data[my_data$EMPSTAT != "NIU (not in universe)" ,]


mean(my_data$AGE) ## 26
mean(sub_data$AGE) ## 32

sub_data <- droplevels(sub_data) ## drop removed labels
```


## Tabulate subset AND store as R object

```{r}
crosstab <- table(sub_data$EMPSTAT, sub_data$SEX)
```


#### print R object
```{r}
crosstab
```


#### write out table(s)
```{r}
write.csv(crosstab,
          file =  "xtab_emp_by_sex.csv")

write.xlsx(crosstab,
           file = "xtab_emp_by_sex.xlsx")

write_dta(as.data.frame(crosstab),
          path = "xtab_emp_by_sex.dta") ## note different syntax
```



## Visualize crosstab

```{r}
barplot(crosstab) ## stacked by default barplot


barplot(crosstab, beside = TRUE) ## or unstacked
```

#### Add a legend

```{r}
par("mar" = c(5.1,4.1,2.1,10.1), xpd = TRUE) ## add space for legend, and turn off clipping

barplot(crosstab, beside = TRUE)
legend("topright",
       legend = levels(sub_data$EMPSTAT),
       pch = 22,
       pt.bg = grey.colors(4),
       inset = c(-1.2,0),
       title = "Employment Status"
       )
```

#### Add some flair

```{r}
par("mar" = c(5.1,4.1,2.1,10.1), xpd = TRUE) ## add space for legend, and turn off clipping

barplot(crosstab,
        beside = TRUE,
        col = rainbow(4),
        main = "Employment by Sex"
        )

legend("topright",
       legend = levels(sub_data$EMPSTAT),
       pch = 22,
       pt.bg = rainbow(4),
       inset = c(-1.2,0),
       title = "Employment Status"
)

```
