---
title: "Example Rmd"
author: "ipumsi"
date: 'July 27th 2022'
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Read in my_data (.csv)

Base R can handle .csv and .txt files "out of the box", but these can have limitations. **Note:** With Rmd, comments/text do not require the "#" comment symbol. Instead we specificy where the `code` will go using "code chunks", as below. 

```{r}

my_data <- read.csv("example_data.csv")

colnames(my_data)
table(my_data$SEX)
table(my_data$EDATTAIN)
table(my_data$EMPSTAT)
```


To read in different file types, we will need to download R packages. 

## Install Packages

R packages are add-ons that expand the functionality of base R. Its helpful  to think of R as a series of books. "Base R" is the starter set of "books" needed for R to function. install.packages() downloads new the "book" and adds it to your collection. This step needs to be done only once, or periodically to update. Note: quotation marks ("") ARE REQUIRED to specify a package name. openxlsx is a very useful package that allows read/write of xls/xlsx files. haven provides functions to read from various stats packages, including SPSS, STATA, and SAS. Notably, haven is able to read in value and value labels

```{r}
if(!require("openxlsx")) {
  install.packages("openxlsx")
}

if (!(require("haven"))) {
  install.packages("haven")
}
```

## Load Packages

In order to load the package, we use library(). This is like taking the book off your shelf and opening it to read/use. This step needs to be done each time you open R/Rstudio. so it's helpful to have a section at the top of the script that loads all packages needed. Note: quotation marks ("") are NOT required for the library function.

```{r}
library(haven)
library(openxlsx)
```

## Read in my_data (.xlsx)



# Read in my_data (.csv) 

Base R can handle .csv and .txt files "out of the box"

Without assigning to an object, contents of file will be PRINTED
```{r}
read.csv("example_data.csv")
```

## read data and ASSIGN to OBJECT
```{r}
my_data_csv <- read.csv("example_data.csv")
```

## View object

Viewing can be done interactively, but note that in order to include the output of `View` in a .Rmd report, we'll need to include a figure.
```{r, eval = FALSE}
View(my_data_cvs)
```

```{r, echo = FALSE}
# knitr::include_graphics()
```

To read in different file types, we will need to download additional R packages.

# Install packages

R packages are add-ons that expand the functionality of base R. Its helpful  to think of R as a series of books. "Base R" is the starter set of "books" needed for R to function. install.packages() downloads new the "book" and adds it to your collection. This step needs to be done only once, or periodically to update. Note: quotation marks ("") ARE REQUIRED to specify a package name. openxlsx is a very useful package that allows read/write of xls/xlsx files. haven provides functions to read from various stats packages, including SPSS, STATA, and SAS. Notably, haven is able to read in value and value labels

**NOTE:** installing packages should be done manually, not as part of an Rmd report.

```{r}
# install.packages("openxlsx")
# install.packages("haven")
# install.packages("ggplot2") ## for plotting
# install.packages("gtsumamry") ## for tables
```
# Load Packages

In order to load the package, we use library(). This is like taking the book off your shelf and opening it to read/use. This step needs to be done each time you open R/Rstudio. so it's helpful to have a section at the top of the script that loads all packages needed. Note: quotation marks ("") are NOT required for the library function.

```{r}
library(haven)
library(openxlsx)
library(ggplot2)
library(gtsummary)
```

# Read in my_data (xlsx)

Excel can be convenient for data management, but like .csv files does not allow for labeled values, like many other stats programs. Coded values may be hard to work with.

```{r}
my_data_xlsx <- read.xlsx("example_data.xlsx")

## Since Viewing doesn't work, we'll comment it out
# View(my_data_xlsx)

#### Summarize (Frequency Tables)
table(my_data_xlsx$SEX)
table(my_data_xlsx$EDATTAIN)
table(my_data_xlsx$EMPSTAT)
```

# Read in my_data (with Labels)

The haven package provides functions to read/write for many common statistical packages. This package makes it easier to get your data into R, and also provides support for LABELED VALUES.

See Environment Panel, haven labels are complex; factors preserve the [value + label] pairs, in more simple structure

We use `as_factor` from the `haven` package to simplify the labeled values.

```{r}
my_data <- read_dta("example_data.dta")


head(my_data) # It's still numeric! and confusing


## as_factor in the haven package to convert for analysis

my_data$SEX <- as_factor(my_data$SEX)
my_data$EDATTAIN <- as_factor(my_data$EDATTAIN)
my_data$EMPSTAT <- as_factor(my_data$EMPSTAT)

head(my_data) # Labels present!
```


## Summaries

#### Summarize (Frequency Tables)
```{r}
table(my_data$SEX)
table(my_data$EDATTAIN)
table(my_data$EMPSTAT)
```
#### Summarize (Numeric)
```{r}
range(my_data$AGE)
```
#### Summarize (Two-Way Tables)
```{r}
table(my_data$EDATTAIN, my_data$SEX)
```
## Basic Data Manipulation 

### Subset by a value

Use `subset()` function specifying a DATA.FRAME followed by a LOGICAL ARGUMENT
```{r}
sub_data <- subset(my_data,
                   EDATTAIN != "Unknown"
                   )

table(sub_data$EDATTAIN) ## empty levels present
```

R factors will "hold onto" empty levels, unless we explicitly tell R to ignore, or "drop" unused levels.
```{r}
sub_data <- droplevels(sub_data) ## drop removed labels


crosstab <- table(sub_data$EDATTAIN, sub_data$SEX)

```

#### A Note on R objects

Calling an R object will print the contents of that object. Small objects may be printed in the console/Rmd report, but larger onces might get truncated:
```{r}
## A Reasonable table
crosstab
```

```{r}
## Large objects may get truncated/summarized
my_data
```


# Write Out R objects

Options to write R objects out to various filetypes. Note the similarities/differences in syntax. In most write- functions, the first argument is the R object to be written out.

```{r}
write.csv(crosstab,
          file =  "xtab_emp_by_sex.csv")

write.xlsx(crosstab,
           file = "xtab_emp_by_sex.xlsx")

write_dta(as.data.frame(crosstab),
          path = "xtab_emp_by_sex.dta") ## note different syntax

```

# Base R vs Packages - Tables

Viewing tables in base R can be a bit clunky. 
```{r}

crosstab ## print frequencies
prop.table(crosstab) ## get proportions

```
#### Fancier Tables (gtsummary)

`gtsummary` produces nice all-in-one tables in the `Viewer` panel, but note the need to get an image to include for reports.

```{r}
library(gtsummary)
tbl_summary(sub_data, by = "SEX")
```

```{r, echo = FALSE}
# knitr::include_graphics()
```
# Base R vs Packages - Graphs

#### Basic Visualization (Base R)

```{r}
barplot(crosstab)
```

Base R provides MANY built in functions, each with many options to customize your plots. The flexibility/customization comes at the trade-off of user input. Adding a legend takes 2 extra steps in Base R.

First, we use layout() to tell R to treat the plot window as a matrix(), consisting of 1 row and 4 columns: 1,1,2,2. The first plot will occupy the first two columns (as a merged column), while the second plot will occupy the last set of columns.

Make an empty plot (of a single point) to use as the base for the legend. "n" indicates "null" for many arguments to the plot function.

```{r}
## plot matrix
layout(matrix(c(1,1,2,2), ncol = 4))

## barplot as normal
barplot(crosstab,
        col = rainbow(4),
        main = "Education by Sex"
)

## Make an empty plot 
plot(1,
     type = "n", ## do not plot the points
     bty = "n", ## do not plot a bounding box
     xaxt = "n", ## do not plot an X axis
     xlab = "", ## label the x axis as "", IE blank
     yaxt = "n", ## do not plot a Y axis
     ylab ="", ## label the Y axis as "", IE blank
     # par("mar") = c(0,0,0,0)
)

legend("center",
       legend = levels(sub_data$EDATTAIN),
       pch = 22,
       pt.bg = rainbow(4),
       inset = c(-1.2,0),
       title = "Ed Attainment"
)
```

#### Basic Visualization (ggplot)

`ggplot` is EXTREMELY popular for quickly generating polished looking plots. It does use a slightly different syntax than Base R, but there is ample user-support.

```{r}

qplot(x = SEX, fill = EDATTAIN, data = sub_data, geom = "bar")

```

